---
title: "README"
output: html_document
---

A script, **run_analysis.R** is designed to perform the tasks listed in the project assignment. The script is 
designed with R version 3.3 and on Windows 8.1 operating systems.

First, the script reads all the relavant files from the working directory and loads them into R. Datasets do not 
have column names, so while loading R provides names ranging from X1, X2, an so on.This part is fairly straight forward, however, the following step needs some additional activities. 

```{r include=FALSE}
require(readr)
require(dplyr)
```


```{r  message = FALSE, warnings = FALSE}
########## Loading the data into R  from working directory ##################
#Get working directory
working_dir <- getwd()

#derive the file paths for the input files

file_path_X_train <- paste(working_dir,"/X_train.txt", sep = "")
file_path_y_train <- paste(working_dir,"/y_train.txt", sep = "")
file_path_subject_train <- paste(working_dir,"/subject_train.txt", sep = "")

file_path_X_test <- paste(working_dir,"/X_test.txt", sep = "")
file_path_y_test <- paste(working_dir,"/y_test.txt", sep = "")
file_path_subject_test <- paste(working_dir,"/subject_test.txt", sep = "")

file_path_activity_labels <- paste(working_dir,"/activity_labels.txt", sep = "")
file_path_features <- paste(working_dir,"/features.txt", sep = "")

 #Training data

        # Load the training data into R
        X_train <- read_table(file_path_X_train,col_names = FALSE)
        
        y_train <- read_table(file_path_y_train,col_names = FALSE)
        
        subject_train <- read_table(file_path_subject_train,col_names = FALSE)
 
 #Test data        
               
        # Load the test data into R
        X_test <- read_table(file_path_X_test ,col_names = FALSE)
        
        y_test <- read_table(file_path_y_test,col_names = FALSE)
        
        subject_test <- read_table(file_path_subject_test ,col_names = FALSE)
        
 #Activity data      
        
        #Load activity data into R
        activity_labels <- read_table(file_path_activity_labels,col_names = FALSE)
        
 #Features data         
        
        #Load features data into R
        features <- invisible(read_csv(file_path_features,col_names = FALSE))
```


In the next step, after analysing the data loaded, it is revealed that features data has the following faults with respect to tidy data :  
    1) column index and the column name both are sitting within one column.  
    2) few columns have same name in common.
    
This data is to be cleaned before processing further, and the code chunk below does the same here :

```{r eval=FALSE}
        # create a table clear_features by removing all the defects mentioned above.
        
        clear_features <- features %>% 
                             separate(X1,c("X1","X2"), sep = " ") %>% # separates column index from column name
                                group_by(X2) %>% 
                                   mutate(cnt = row_number(X2)) %>% 
                                      unite( X3, X2,cnt)              #removes the duplicate names by appending
                                                                      #occurance at the end and stores them in 
                                                                      #column X3
```


Next thing is to be done is to join the activities with theirs lables.

```{r eval=FALSE}
########## Joining of activities with its label ##################        
                        
          # y_train contains training activities corresponds to x_train data set.
          #However, its labels are there in the activitiy_lables data set.
                                  
          # create a data set training_activity_lables to contain activities and their descriptions 
                                  
          training_activity_lables <- left_join(y_train,activity_labels, by = "X1")  
          
          #similarly, create a data set test_activity_lables to contain activities and their descriptions 
          test_activity_lables <- left_join(y_test,activity_labels, by = "X1")  
```

Now everything is ready to go ahead with the tasks. Here I take little liberty to change the order of the tasks
for my convenience. The order is as follows :

* Task 4  
* Task 2  
* Task 3  
* Task 1  
* Task 5

Starting with task 4 first.

```{r eval=FALSE}
########## TASK 4 ##################
#Appropriately labels the data set with descriptive variable names.

          #Label the training data with appropriate labels
          
          names(X_train) <- clear_features$X3 
          
          #Label the test data with appropriate labels
          
          names(X_test) <- clear_features$X3 
```

Task 2 is about the selecting the measurements related to mean and std.

```{r eval=FALSE}
########## TASK 2 ##################

#Extracts only the measurements on the mean and standard deviation for each measurement.

          # extract training data and store it to data_extract_training
          data_extract_training <- select(X_train, matches("mean|std"))
          
          # extract test data and store it to data_extract_test
          data_extract_test <- select(X_test, matches("mean|std"))
```

Task 3 is all about adding descriptive information for activities and subjects.

```{r eval=FALSE}
########## TASK 3 ##################

#Uses descriptive activity names to name the activities in the data set
          
    #Training data
        
        #Binding activities    

          #Add descriptve activity names to the training dataset and rename it to 'activity'
          
          data_extract_training <- bind_cols(as.data.frame(training_activity_lables$X2),data_extract_training)
          
          names(data_extract_training)[1] <- 'activity'
          
        #Binding subjects     
          
          #Add subject number to the training dataset and rename it to 'subject'
          
          data_extract_training <- bind_cols(as.data.frame(subject_train$X1),data_extract_training)
          
          names(data_extract_training)[1] <- 'subject'
          
    #Test data  
          
        #Binding activities       
          
          #Add descriptve activity names to the test dataset and rename it to 'activity'
          
          data_extract_test <- bind_cols(as.data.frame(test_activity_lables$X2),data_extract_test)
          
          names(data_extract_test)[1] <- 'activity'
          
        #Binding subjects     
          
          #Add subject number to the training dataset and rename it to 'subject'
          
          data_extract_test <- bind_cols(as.data.frame(subject_test$X1),data_extract_test)
          
          names(data_extract_test)[1] <- 'subject'
```

And now the Task 1 to merge both the datasets test and training to a single one.

```{r eval=FALSE}
########## TASK 1 ##################
#Merges the training and the test sets to create one data set.

          #create tidy_data_extract by merging data_extract_test and data_extract_training.
          
          tidy_data_extract <- bind_rows(data_extract_test,data_extract_training)
```

Once the above steps are executed, data is ready to perform the final task, Task 5, which is about to create
a seperate data  set with the average values for each activity anf subject.

```{r eval=FALSE}
########## TASK 5 ##################
#From the data set in step 4, creates a second, independent tidy data set with the average of 
          #each variable for each activity and each subject.    
          
          # perform the operation and store it into each_val_avg
          each_val_avg <- tidy_data_extract %>%
                             group_by(activity,subject) %>%
                                summarise_each(funs(mean))
```

The final dataset which is stored into each_val_avg is having the following details.

```{r eval=FALSE}
Classes 'grouped_df', 'tbl_df', 'tbl' and 'data.frame':	180 obs. of  85 variables:
 $ activity                         : Factor w/ 6 levels "LAYING","SITTING",..: 1 1 1 1 1 1 1 1 1 1 ...
 $ subject                          : int  1 2 3 4 5 6 7 8 9 10 ...
 $ tBodyAcc-mean()-X_1              : num  0.222 0.281 0.276 0.264 0.278 ...
 $ tBodyAcc-mean()-Y_1              : num  -0.0405 -0.0182 -0.019 -0.015 -0.0183 ...
 $ tBodyAcc-mean()-Z_1              : num  -0.113 -0.107 -0.101 -0.111 -0.108 ...
 $ tBodyAcc-std()-X_1               : num  -0.928 -0.974 -0.983 -0.954 -0.966 ...
 $ tBodyAcc-std()-Y_1               : num  -0.837 -0.98 -0.962 -0.942 -0.969 ...
 $ tBodyAcc-std()-Z_1               : num  -0.826 -0.984 -0.964 -0.963 -0.969 ...
 $ tGravityAcc-mean()-X_1           : num  -0.249 -0.51 -0.242 -0.421 -0.483 ...
 $ tGravityAcc-mean()-Y_1           : num  0.706 0.753 0.837 0.915 0.955 ...
 $ tGravityAcc-mean()-Z_1           : num  0.446 0.647 0.489 0.342 0.264 ...
 $ tGravityAcc-std()-X_1            : num  -0.897 -0.959 -0.983 -0.921 -0.946 ...
 $ tGravityAcc-std()-Y_1            : num  -0.908 -0.988 -0.981 -0.97 -0.986 ...
 $ tGravityAcc-std()-Z_1            : num  -0.852 -0.984 -0.965 -0.976 -0.977 ...
 $ tBodyAccJerk-mean()-X_1          : num  0.0811 0.0826 0.077 0.0934 0.0848 ...
 $ tBodyAccJerk-mean()-Y_1          : num  0.00384 0.01225 0.0138 0.00693 0.00747 ...
 $ tBodyAccJerk-mean()-Z_1          : num  0.01083 -0.0018 -0.00436 -0.00641 -0.00304 ...
 $ tBodyAccJerk-std()-X_1           : num  -0.958 -0.986 -0.981 -0.978 -0.983 ...
 $ tBodyAccJerk-std()-Y_1           : num  -0.924 -0.983 -0.969 -0.942 -0.965 ...
 $ tBodyAccJerk-std()-Z_1           : num  -0.955 -0.988 -0.982 -0.979 -0.985 ...
 $ tBodyGyro-mean()-X_1             : num  -0.01655 -0.01848 -0.02082 -0.00923 -0.02189 ...
 $ tBodyGyro-mean()-Y_1             : num  -0.0645 -0.1118 -0.0719 -0.093 -0.0799 ...
 $ tBodyGyro-mean()-Z_1             : num  0.149 0.145 0.138 0.17 0.16 ...
 $ tBodyGyro-std()-X_1              : num  -0.874 -0.988 -0.975 -0.973 -0.979 ...
 $ tBodyGyro-std()-Y_1              : num  -0.951 -0.982 -0.977 -0.961 -0.977 ...
 $ tBodyGyro-std()-Z_1              : num  -0.908 -0.96 -0.964 -0.962 -0.961 ...
 $ tBodyGyroJerk-mean()-X_1         : num  -0.107 -0.102 -0.1 -0.105 -0.102 ...
 $ tBodyGyroJerk-mean()-Y_1         : num  -0.0415 -0.0359 -0.039 -0.0381 -0.0404 ...
 $ tBodyGyroJerk-mean()-Z_1         : num  -0.0741 -0.0702 -0.0687 -0.0712 -0.0708 ...
 $ tBodyGyroJerk-std()-X_1          : num  -0.919 -0.993 -0.98 -0.975 -0.983 ...
 $ tBodyGyroJerk-std()-Y_1          : num  -0.968 -0.99 -0.987 -0.987 -0.984 ...
 $ tBodyGyroJerk-std()-Z_1          : num  -0.958 -0.988 -0.983 -0.984 -0.99 ...
 $ tBodyAccMag-mean()_1             : num  -0.842 -0.977 -0.973 -0.955 -0.967 ...
 $ tBodyAccMag-std()_1              : num  -0.795 -0.973 -0.964 -0.931 -0.959 ...
 $ tGravityAccMag-mean()_1          : num  -0.842 -0.977 -0.973 -0.955 -0.967 ...
 $ tGravityAccMag-std()_1           : num  -0.795 -0.973 -0.964 -0.931 -0.959 ...
 $ tBodyAccJerkMag-mean()_1         : num  -0.954 -0.988 -0.979 -0.97 -0.98 ...
 $ tBodyAccJerkMag-std()_1          : num  -0.928 -0.986 -0.976 -0.961 -0.977 ...
 $ tBodyGyroMag-mean()_1            : num  -0.875 -0.95 -0.952 -0.93 -0.947 ...
 $ tBodyGyroMag-std()_1             : num  -0.819 -0.961 -0.954 -0.947 -0.958 ...
 $ tBodyGyroJerkMag-mean()_1        : num  -0.963 -0.992 -0.987 -0.985 -0.986 ...
 $ tBodyGyroJerkMag-std()_1         : num  -0.936 -0.99 -0.983 -0.983 -0.984 ...
 $ fBodyAcc-mean()-X_1              : num  -0.939 -0.977 -0.981 -0.959 -0.969 ...
 $ fBodyAcc-mean()-Y_1              : num  -0.867 -0.98 -0.961 -0.939 -0.965 ...
 $ fBodyAcc-mean()-Z_1              : num  -0.883 -0.984 -0.968 -0.968 -0.977 ...
 $ fBodyAcc-std()-X_1               : num  -0.924 -0.973 -0.984 -0.952 -0.965 ...
 $ fBodyAcc-std()-Y_1               : num  -0.834 -0.981 -0.964 -0.946 -0.973 ...
 $ fBodyAcc-std()-Z_1               : num  -0.813 -0.985 -0.963 -0.962 -0.966 ...
 $ fBodyAcc-meanFreq()-X_1          : num  -0.159 -0.146 -0.074 -0.274 -0.136 ...
 $ fBodyAcc-meanFreq()-Y_1          : num  0.0975 0.2573 0.2385 0.3662 0.4665 ...
 $ fBodyAcc-meanFreq()-Z_1          : num  0.0894 0.4025 0.217 0.2013 0.1323 ...
 $ fBodyAccJerk-mean()-X_1          : num  -0.957 -0.986 -0.981 -0.979 -0.983 ...
 $ fBodyAccJerk-mean()-Y_1          : num  -0.922 -0.983 -0.969 -0.944 -0.965 ...
 $ fBodyAccJerk-mean()-Z_1          : num  -0.948 -0.986 -0.979 -0.975 -0.983 ...
 $ fBodyAccJerk-std()-X_1           : num  -0.964 -0.987 -0.983 -0.98 -0.986 ...
 $ fBodyAccJerk-std()-Y_1           : num  -0.932 -0.985 -0.971 -0.944 -0.966 ...
 $ fBodyAccJerk-std()-Z_1           : num  -0.961 -0.989 -0.984 -0.98 -0.986 ...
 $ fBodyAccJerk-meanFreq()-X_1      : num  0.132 0.16 0.176 0.182 0.24 ...
 $ fBodyAccJerk-meanFreq()-Y_1      : num  0.0245 0.1212 -0.0132 0.0987 0.1957 ...
 $ fBodyAccJerk-meanFreq()-Z_1      : num  0.0244 0.1906 0.0448 0.077 0.0917 ...
 $ fBodyGyro-mean()-X_1             : num  -0.85 -0.986 -0.97 -0.967 -0.976 ...
 $ fBodyGyro-mean()-Y_1             : num  -0.952 -0.983 -0.978 -0.972 -0.978 ...
 $ fBodyGyro-mean()-Z_1             : num  -0.909 -0.963 -0.962 -0.961 -0.963 ...
 $ fBodyGyro-std()-X_1              : num  -0.882 -0.989 -0.976 -0.975 -0.981 ...
 $ fBodyGyro-std()-Y_1              : num  -0.951 -0.982 -0.977 -0.956 -0.977 ...
 $ fBodyGyro-std()-Z_1              : num  -0.917 -0.963 -0.967 -0.966 -0.963 ...
 $ fBodyGyro-meanFreq()-X_1         : num  -0.00355 0.10261 -0.08222 -0.06609 -0.02272 ...
 $ fBodyGyro-meanFreq()-Y_1         : num  -0.0915 0.0423 -0.0267 -0.5269 0.0681 ...
 $ fBodyGyro-meanFreq()-Z_1         : num  0.0105 0.0553 0.1477 0.1529 0.0414 ...
 $ fBodyAccMag-mean()_1             : num  -0.862 -0.975 -0.966 -0.939 -0.962 ...
 $ fBodyAccMag-std()_1              : num  -0.798 -0.975 -0.968 -0.937 -0.963 ...
 $ fBodyAccMag-meanFreq()_1         : num  0.0864 0.2663 0.237 0.2417 0.292 ...
 $ fBodyBodyAccJerkMag-mean()_1     : num  -0.933 -0.985 -0.976 -0.962 -0.977 ...
 $ fBodyBodyAccJerkMag-std()_1      : num  -0.922 -0.985 -0.975 -0.958 -0.976 ...
 $ fBodyBodyAccJerkMag-meanFreq()_1 : num  0.266 0.342 0.239 0.274 0.197 ...
 $ fBodyBodyGyroMag-mean()_1        : num  -0.862 -0.972 -0.965 -0.962 -0.968 ...
 $ fBodyBodyGyroMag-std()_1         : num  -0.824 -0.961 -0.955 -0.947 -0.959 ...
 $ fBodyBodyGyroMag-meanFreq()_1    : num  -0.1398 0.0186 -0.0229 -0.2599 0.1024 ...
 $ fBodyBodyGyroJerkMag-mean()_1    : num  -0.942 -0.99 -0.984 -0.984 -0.985 ...
 $ fBodyBodyGyroJerkMag-std()_1     : num  -0.933 -0.989 -0.983 -0.983 -0.983 ...
 $ fBodyBodyGyroJerkMag-meanFreq()_1: num  0.1765 0.2648 0.1107 0.2029 0.0247 ...
 $ angle(tBodyAccMean_1             : num  0.02137 0.00579 0.0179 -0.00236 0.02121 ...
 $ angle(tBodyAccJerkMean)_1        : num  0.00306 -0.00636 0.01665 -0.0155 0.05536 ...
 $ angle(tBodyGyroMean_1            : num  -0.00167 0.06529 0.04559 0.00713 -0.00581 ...
 $ angle(tBodyGyroJerkMean_1        : num  0.0844 -0.0294 0.0317 0.0367 -0.0344 ...
```

Finally, write the dataset into a text file - tidy_data_set.txt.

```{r eval=FALSE}
########## WRITING THE OUTPUT TO A FILE ##################
file_path_output <- paste(working_dir,"/tidy_data_set.txt", sep = "")
          
          write.table(each_val_avg, file = file_path_output, row.names = FALSE)
          
```


## Output

After running the script **run_analysis.R** a text file **tidy_data_set.txt** will be generated at the working directory.